Storm surge modelling in the Strait of Georgia

### 2014-11-25 RoughWork/manuscript.tex
Comments on local winds/eastern Canada

Communities in the Strait of Georgia are particularly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal ranges occur in June and December at the summer and winter solstices \citep{thomson1981oceanography}. The highest water levels are observed in the winter around the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. As such, storm surges in the Strait of Georgia are typically observed between November and February \citep{forseth2006adaptation}.

@@

Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} in a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters by \citet{murty1995storm} and \citet{danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting}, relatively little has been published on storm surge modelling on the west coast of Canada.

@@

Hindcasts of four significant events in February 2006, November 2006, December 2006 and November 2009 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude ($>50$\%), in agreement with previous studies in this region \citep{murty1995storm}. We have also found that the wind and atmospheric conditions can set up spatial differences in the surge amplitude, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important after the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected.

Comparisons between high resolution and low resolution atmospheric forcing further suggest that the local winds are not the leading order factor in setting up large surge amplitudes. The low resolution model significantly underestimated wind speeds in certain areas of the domain, however, it performed just as accurately as the high resolution product.  Although the high resolution model delivers fine scale detail and more accurate local wind patterns, it did not provide any improvement to the surge hindcast. Since both of these products provide the same level of accuracy in storm surge hindcasts, in the future we will use the high resolution atmospheric model to produce real-time storm surge predictions, even though most of out testing was performed with the low resolution model.

>>>>

Communities in the Strait of Georgia are particularly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal ranges occur in June and December at the summer and winter solstices \citep{thomson1981oceanography}. The highest water levels are observed in the winter around the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. As such, storm surges in the Strait of Georgia are typically observed between November and February \citep{forseth2006adaptation}. Flooding caused by a storm surge is most likely to occur during a high spring tide \citep{abeys2011extreme}.

@@

Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} in a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. A statisical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm ENSO periods \citep{abeys2011extreme}.

@@

Hindcasts of five significant events in February 2006, November 2006, December, November 2009, and December 2012 were simulated. In general, the model performs well at reproducing the observed water level heights and surge amplitudes, with Willmott skill scores greater than 0.88 in all cases. We have also assessed the relative importance of different forcing factors contributing to the storm surge. The sea surface height anomaly entering the system from the Pacific Ocean contributes most significantly to the surge amplitude, in agreement with previous studies in this region \citep{murty1995storm}. Although the large scale wind patterns and atmopsheric conditions important in setting up the surge on the Pacific Ocean boundaries, the local winds within the Strait of Georgia are not the leading order contribution to surge amplitude. However, the local wind patterns can set up small spatial differences in the surge amplitude, resulting in higher surges on the mainland side of the Strait of Georgia due to the usually southerly winds during storm surge season. Additionally, surge amplitudes in the Strait of Georgia are typically higher than those in the Strait of Juan de Fuca, perhaps due to anomalies entering from both the northern and southern regions. Results from an additional simulation with no tidal forcing but only surge propagation and atmospheric forcing found that nonlinear effects due to the tide-surge interaction were most important after the San Juan and Gulf Islands. The modelled surge amplitude increased by about 10\% at Point Atkinson when the tide-surge interaction was neglected.

Comparisons between high resolution and low resolution atmospheric forcing further suggest that the local winds are not the leading order factor in setting up large surge amplitudes. The low resolution model significantly underestimated wind speeds in certain areas of the domain, however, it performed just as accurately in reproducing surge amplitudes as the high resolution product.  Although the high resolution model delivers fine scale detail and more accurate local wind patterns, it did not provide any improvement to the surge hindcast. Since both of these products provide the same level of accuracy in storm surge hindcasts, in the future we will use the high resolution atmospheric model to produce real-time storm surge predictions, even though most of out testing was performed with the low resolution model.
